Allocation Removal by Partial Evaluation in a Tracing JIT

### 2010-10-14 talk/pepm2011/paper.tex
fixes by david

The goal of a just-in-time (JIT) compiler for a dynamic language is obviously to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is thus to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing of primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of the interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many optimizations that are usually
hard in a compiler, such as register allocation.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
thus mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

This trace of operations is then the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation is simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through it.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \lstinline{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \lstinline{if} statement is true, because if
it isn't, the rest of the trace is not valid.

@@

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long -- they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is thus to analyze which objects fall in category 1 and thus do
not have to be allocated at all.

@@

The static heap is a partial function from $V^*$ into the
set of static objects, which are triples of a type and two elements of $V^*$.
A variable $v^*$ is in the domain of the static heap $S$ as long as the
optimizer can fully keep track of the object. The image of $v^*$ is what is
statically known about the object stored in it, \ie its type and its fields. The
fields of objects in the static heap are also elements of $V^*$ (or null, for
short periods of time).

@@

In addition to the count of operations we also performed time measurements.
All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{georges_statistically_2007}. The results are reported in
Figure~\ref{fig:times}. In addition to the run times the table also reports the
speedup that PyPy with optimization turned on achieves.
With the optimization turned on, PyPy's Python interpreter outperforms CPython
in all benchmarks except spambayes (which heavily relies on regular expression
performance and thus is not helped much by our Python JIT) and meteor-contest.
All benchmarks are improved by the allocation removal optimization, by at least
20\% and by as much as a factor of 6.95.

@@

There exists a large number of works on escape analysis, which is an program
analysis that tries to find an upper bound for the lifetime of objects allocated
at specific program points
\cite{goldberg_higher_1990,park_escape_1992,choi_escape_1999,bruno_blanchet_escape_2003}.
This information can then be used to decide that certain objects can be
allocated on the stack, because their lifetime does not exceed that of the stack
frame it is allocated in. The difference to our work is that escape analysis is
split into an analysis and an optimization phase. The analysis can be a lot more
complex than our simple one-pass optimization. Also, stack-allocation reduces
garbage-collection pressure but does not optimize away the actual accesses to
the stack-allocated object. In our case, an object is not needed at all any
more.
XXX more papers?

@@

In this paper, we used an approach based on online partial evaluation
to optimize away allocations and type guards in the traces of a
tracing JIT.  In this context a simple approach to partial evaluation
gives good results.  This is due to the fact that the tracing JIT
itself is responsible for all control issues, which are usually the
hardest part of partial evaluation: the tracing JIT selects the parts
of the program that are worthwhile to optimize, and extracts linear
paths through them, inlining functions as necessary.  What is left to
optimize is only those linear paths.

>>>>

The goal of a just-in-time (JIT) compiler for a dynamic language is obviously to
improve the speed of the language over an implementation of the language that
uses interpretation. The first goal of a JIT is therefore to remove the
interpretation overhead, i.e. the overhead of bytecode (or AST) dispatch and the
overhead of the interpreter's data structures, such as operand stack etc. The
second important problem that any JIT for a dynamic language needs to solve is
how to deal with the overhead of boxing primitive types and of type
dispatching. Those are problems that are usually not present or at least less
severe in statically typed languages.

@@

A recently popular approach to implementing just-in-time compilers for dynamic
languages is that of a tracing JIT. A tracing JIT works by observing the running
program and recording its hot spots into linear execution traces, which are then turned into
machine code. One reason for the popularity of tracing JITs is their relative
simplicity. They can often be added to an interpreter and a lot of the
infrastructure of the interpreter can be reused. They give some important
optimizations like inlining and constant-folding for free. A tracing JIT always
produces linear pieces of code, which simplifies many algorithms that are usually
hard in a compiler, such as register allocation.

@@

The core idea of tracing JITs is to focus the optimization effort of the JIT
compiler on the hot paths of the core loops of the program and to just use an
interpreter for the less commonly executed parts. VMs that use a tracing JIT are
mostly mixed-mode execution environments, they contain both an interpreter and a
JIT compiler. By default the interpreter is used to execute the program, doing
some light-weight profiling at the same time. This profiling is used to identify
the hot loops of the program. If a hot loop is found in that way, the
interpreter enters a special \emph{tracing mode}. In this tracing mode, the
interpreter tries to record all operations that it is executing while running one
iteration of the hot loop. This history of executed operations of one loop is
called a \emph{trace}. Because the trace corresponds to one iteration of a loop,
it always ends with a jump to its own beginning. The trace also contains all
operations that are performed in functions that were called in the loop, thus a
tracing JIT automatically performs inlining.

This trace of operations is then the basis of the generated code. The trace is
first optimized, and then turned into machine code. Both optimization
and machine code generation are simple, because the traces are linear. This
linearity makes many optimizations a lot more tractable, and the inlining that
happens gives the optimizations automatically more context to work with.

Since the trace corresponds to one concrete execution of a loop,
the code generated from it is only one possible path through it.
To make sure that the trace is maintaining the correct semantics, it contains a
\emph{guard} at all places where the execution could have diverged from the
path. Those guards check the assumptions under which execution can stay on the
trace. As an example, if a loop contains an \lstinline{if} statement, the trace
will contain the execution of one of the paths only, which is the path that was
taken during the production of the trace. The trace will also contain a guard
that checks that the condition of the \lstinline{if} statement is the same as
during tracing, because if
it isn't, the rest of the trace is not valid. \cfbolz{The "if" shouldn't be bold}

@@

The main insight to improve the code shown in the last section is that objects
in category 1 don't survive very long -- they are used only inside the loop and
nobody else in the program stores a reference to them. The idea for improving
the code is to analyze which objects fall in category 1 and therefore do
not have to be allocated at all.

@@

The static heap is a partial function from $V^*$ into the
set of static objects, which are triples of a type and two elements of $V^*$.
A variable $v^*$ is in the domain of the static heap $S$ as long as the
optimizer can fully keep track of the object. The object $S(v^*)$ is what is
statically known about the object stored in it, \ie its type and its fields. The
fields of objects in the static heap are also elements of $V^*$ (or null, for
short periods of time).

@@

In addition to the count of operations we also performed time measurements.
All benchmarks were run 50 times in the same process, to give the JIT time to produce machine
code. The arithmetic mean of the times of the last 30 runs were used as the
result. The errors were computed using a confidence interval with a 95\%
confidence level \cite{georges_statistically_2007}. The results are reported in
Figure~\ref{fig:times}. In addition to the run times the table also reports the
speedup that PyPy achieves when the optimization is turned on.

With the optimization turned on, PyPy's Python interpreter outperforms CPython
in all benchmarks except spambayes (which heavily relies on regular expression
performance and thus is not helped much by our Python JIT) and meteor-contest.
All benchmarks are improved by the allocation removal optimization, by at least
20\% and by as much as a factor of 6.95.

@@

There exists a large number of works on escape analysis, which is a program
analysis that tries to find an upper bound for the lifetime of objects allocated
at specific program points
\cite{goldberg_higher_1990,park_escape_1992,choi_escape_1999,bruno_blanchet_escape_2003}.
This information can then be used to decide that certain objects can be
allocated on the stack, because their lifetime does not exceed that of the stack
frame it is allocated in. The difference to our work is that escape analysis is
split into an analysis and an optimization phase. The analysis can be a lot more
complex than our simple one-pass optimization. Also, stack-allocation reduces
garbage-collection pressure but does not optimize away the actual accesses to
the stack-allocated object. In our case, an object is not needed at all any
more.
XXX more papers?

@@

In this paper, we used an approach based on online partial evaluation
to optimize away allocations and type guards in the traces of a
tracing JIT.  In this context a simple approach to partial evaluation
gives good results.  This is due to the fact that the tracing JIT
itself is responsible for all control issues, which are usually the
hardest part of partial evaluation: the tracing JIT selects the parts
of the program that are worthwhile to optimize, and extracts linear
paths through them, inlining functions as necessary.  What is left to
optimize are only those linear paths.
